Exploitation Techniques and Mitigations

### 2016-01-24 docs/paper.tex
Paper: Review ROP, ASLR and StackGuard

The first mitigation technique discussed, DEP, is also known under the term
\textit{write XOR execute (\texttt{w\^{}x})} and it will prevent us from
executing injected code like we did in the shell code example previously. This
happens by attaching a execute flag to each page (in addition to the read /
write flags). If the instruction pointer points to a page without the execute
flag set, a segmentation fault will be triggered. The only pages flagged for
execution are the ones that belong either to the \texttt{.text} segment or the
used libraries since it will contain the program code. The stack is (of course)
not executable per default. So our shell code example would simply segfault.

On modern platforms this is enforced by hardware, but even CPUs that lack such
hardware support, software-enforced DEP provides limited
protection.~\cite{wiki:dep}

Previously we worked around this by passing \texttt{-z execstack}, hence data
on the stack was executable. This can also be seen in the output of
\texttt{readelf}.

@@

\Cref{src:dep_readelf} shows that with \texttt{-z execstack} the stack section
is marked (looking at \texttt{Flg} in lines 8 and 19) with execute
(\texttt{E}). Running the example, see \cref{src:dep_shell_code}, yields the
segfault described in the previous paragraph. The program is terminated upon
receiving the segfault.

DEP makes injection arbitrary code into binary much harder, but we still
control the return address.

@@

The first example shown in the Buffer Overflow section
(\cref{sec:buffer_overflow}) already illustrated the power that comes along
with controlling the return address. In there we were able to jump to a
completely different function upon execution. Most likely the target binary may
not contain a function doing exactly what an attacker wants to do, by
controlling the return address one can build a, so called, \textit{ROP chain}
to execute (more or less) arbitrary code, which is made out of
\textit{gadgets}.~\cite{wiki:rop,buchanan2008good,shacham2007}

@@

A gadget is a (short) sequence of instructions ending with a \texttt{ret}.

An attacker scavenges the target binary (and used libraries) for such sequences
in order to combine them to build a malicious sequence of instructions. The
size and diversity of the code base dictates the diversity of available gadgets
and therefore the difficulty of building a specific sequence.

The shell code used in this paper could be mapped to three gadgets, for
example, if the described gadgets were available. The first one pushes the
string \texttt{"/bin//sh"} onto the stack, the second one sets up the registers
(arguments) and the last one calls the \texttt{execve} system call.

The final \texttt{ret} statement of a gadgets is required for chaining them
together. We not only use the Buffer Overflow to control one, but multiple
return addresses. Because of this our payload will be made up by a list of
return addresses (the start of each gadgets) interleaved by some padding. Each
provided address will be consumed one by one at the end of each gadgets to
\emph{jump} to the next one.

@@

The easiest way to find what gadgets are available may be by piping the output
of \texttt{objdump} to \texttt{grep} filtering for \texttt{ret} statements.
This is done in \cref{src:rop_gadgets}. In there you can see three possible
gadgets with different instructions. Gadgets can of course be arbitrary long,
we just used a length of 5 instructions for this example.

@@

\begin{listing}[h!]
    \cfile{../src/return_oriented_programming/rop.c}
    \caption{Example for exploiting a Buffer Overflow with ROP}
    \label{src:rop_example}
\end{listing}

@@

% TODO

@@

ASLR is enabled by default and one can easily check the implications by running
\texttt{cat} on \texttt{/proc/self/maps} a few times as shown in
\cref{src:aslr_cat}. Line 10, 19 and 28 show, for example, that the stack
starts at different locations in memory each time cat has been invoked.

We can directly see one flaw in this setup --- not all sections of the
\texttt{cat} binary start at random locations. Especially the \texttt{.text}
always starts at the same position. This happens because \texttt{cat} itself
was not compiled as Position Independent Executable (PIE). Since this is
actually the default behaviour, most programs' \texttt{.text} segment will
always start at the same location. One could pass the corresponding flag
(\texttt{-pie}) to the compiler to prevent this, then ASLR would be able to
randomize these segments too.

Breaking ASLR, even when the code is compiled with \texttt{-pie}, is easier
than it seems at first. Relocation only happens to a section at whole, inside a
section still shares the same relative distance as it would without ASLR. But
before exploit this fact have a look at the randomize addresses again.

Only three nybble ($3 \times \SI{4}{\bit}$) differ between multiple runs giving
us $2^{12} = 4096$ possibilities. If the scenario allows it, brute forcing is a
viable option here. But note that this changes drastically for \SI{64}{\bit}.

@@

Nevertheless a better option has already been teased. ASLR can be broken easily
as soon as we get \emph{one} address (pointer) to a section of interest is
\emph{leaked}. Therefore the name information leak. We show the implications of
such a leak by an example taken from \cite{rpisec}.

@@

In case you may wonder how easy it is to leak information like a pointer, this
already happened to us as side effect back at the format string example
(\cref{sec:format_string_exploit}).

The adaptation for the exploit we were using looks as follows. First, manage to
leak a pointer somehow which enables you to calculate the address offset
introduced by ASLR. Augment your ROP chain to take the offset into account. Run
the exploit. Since this rather simple and we already gave an example how to
calculate the offset, we would like to leave this as an exercise for the
reader.

Adapt the target file used in the ROP example by printing the address of
\texttt{printf} first and then read in the payload via \texttt{stdin}. This way
you can first simulate a leaked pointer, adapt the ROP chain and run it. Double
check the distances between functions in the libraries, they may not correlate
with the ones used in this writeup.

@@

DEP can be fooled by ROP and ASLR is rendered useless with a simple info leak.
Something else is required at this point. Thinking back, the original
vulnerability was the overwrite of the return address located on the stack. Two
(additional) counter mechanisms were introduced going by the names StackGuard
and StackShield. We will take only a look at StackGuard and one relative common
scenario to break it, but there is a comprehensive article
\cite{phrack_stack_guard} on \texttt{phrack}\footnote{\url{http://phrack.org/}}
describing and breaking both techniques.

The general idea behind StackGuard is to place \emph{something} before the
return address which \emph{guards} against buffer overflows. This
\emph{something} known as a Canary and comes in different forms.

\begin{description}
    \item[Terminator] A terminator Canary contains a sequence of commonly used
        terminator symbols (like null, EOF, linefeed, \dots) to
        \emph{terminate} most string operations. Hence the return address will
        not be overwritten.
    \item[Random] A random Canary is chosen at program start and stored
        \textit{somewhere save} and pushed into the stack upon function calls
        to \emph{guard} the return address. The canary on the stack is checked
        against the one stored save before issuing the return. The program is
        terminated upon mismatch. With this setup an attacker has to know the
        canary in order to overwrite the return address. Since it is chosen at
        random when the program starts, he cannot reliably reproduce the same
        canary in his cloned setup.

@@

There is also the random XOR Canary which XORs the random Canary with the
return address before placing it on the stack. ``This is effectively encryption
of the return address with the random canary of this
function.''~\cite{phrack_stack_guard}

@@

As can be seen in \cref{src:canary_gcc} the buffer was not filled beyond his
capacity to examine the Canary located in the same stack frame. A script
created by Daniel Walter\footnote{\url{http://0x90.at/post/gdb-stack-script}}
has been adapted slightly to display the stack together with some annotations.
Using \texttt{"AAAAAAA\textbackslash{}n"} and
\texttt{"BBBBBBB\textbackslash{}n"} makes the buffer clearly visible in lines
16 and 17. The Canary can be observed in line 18.

The Canary itself is composed of a terminator (null) followed by a random
sequence of \SI{3}{\byte}. This sequence changes very time the program is run.
Feeding more data to the buffer and overflowing it this way yields termination
of the program. Note that the \texttt{puts} function is still called, the
termination happens just before the return of the function.

\begin{code*}{linenos=false}
    > ./vuln <<< "AAAAAAAADEADBEEF"
    AAAAAAAADEADBEEF

@@

>>>>

The first mitigation technique discussed, DEP, is also known under the term
\textit{write XOR execute (\texttt{w\^{}x})} and it will prevent us from
executing injected code like we did in the shell code example previously. This
happens by attaching an execute flag to each page (in addition to the read /
write flags). If the instruction pointer points to a page without the execute
flag set, a segmentation fault will be triggered. The only pages flagged for
execution are the ones that belong either to the \texttt{.text} segment or to
used libraries (by default) since they will contain the program code. The stack
is (of course) not executable by default. Therefore our shell code example
would simply segfault.

The execution flag is enforced by hardware on modern platforms. But for CPUs
that lack such hardware support, software-enforced DEP provides limited
protection.~\cite{wiki:dep}

Previously we worked around this by passing \texttt{-z execstack} to the
compiler, hence data on the stack was executable. This can also be seen in the
output of \texttt{readelf}.

@@

\Cref{src:dep_readelf} shows that with \texttt{-z execstack} the stack section
is marked (looking at \texttt{Flg} in lines 8 and 19) with execute
(\texttt{E}). Running the example, see \cref{src:dep_shell_code}, yields the
segfault described in the previous paragraphs. The program is terminated upon
receiving the segfault.

DEP makes injection arbitrary code into binary much harder, but we still
control the return address --- so let's use it.

@@

The first example shown in the \cref{sec:buffer_overflow} already illustrated
the power that comes along with controlling the return address. It enabled us
to jump to a completely different function upon execution. The target binary
may not contain a function doing exactly what an attacker wants to do. But by
controlling the return address one can build a, so called, \textit{ROP chain}
to execute (more or less) arbitrary code, which is made out of
\textit{gadgets}.~\cite{wiki:rop,buchanan2008good,shacham2007}

@@

A gadget is a, usually short, sequence of instructions ending with a
\texttt{ret} instruction.

An attacker scavenges the target binary (and used libraries) for such sequences
in order to combine them to build a new, malicious sequence of instructions.
The size and diversity of the code base dictates the diversity of available
gadgets and therefore the difficulty of building a specific sequence.

The shell code used in this paper could be mapped to three gadgets, for
example. The first one pushes the string \texttt{"/bin//sh"} onto the stack,
the second one sets up the registers (arguments) and the last one calls the
\texttt{execve} system call.

The final \texttt{ret} statement of a gadgets is required for chaining them
together. We not only use the Buffer Overflow to control \emph{one}, but
\emph{multiple} return addresses. Because of this our payload will be made up
by a list of return addresses (the start of each gadgets) interleaved by some
padding. Each provided address will be consumed one by one at the end of each
gadgets to \emph{jump} to the next one. The payload may still contain data like
\texttt{"/bin//sh"} if necessary.

@@

One can easily get a list of available gadgets by piping the output of
\texttt{objdump} to \texttt{grep} filtering for \texttt{ret} instructions. This
is done in \cref{src:rop_gadgets} where three different gadgets can be
observed. Of course each gadget can be arbitrary long, we just used a length of
5 instructions in this example.

@@

\begin{listing}[h!]
    \begin{minipage}[t]{0.4\textwidth}
        \cfile{../src/return_oriented_programming/rop.c}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.6\textwidth}
        \codefile[breaklines]{../src/return_oriented_programming/run.txt}
    \end{minipage}
    \caption{Example for exploiting a Buffer Overflow with ROP}
    \label{src:rop_example}
\end{listing}

@@

On the right hand side we see the execution of the exploit. First note that we
no longer compile the binary with \texttt{-z execstack}. We read the locations
of \texttt{not\_used} and \texttt{system} via \texttt{gdb} and note down the
corresponding addresses. \texttt{objdump} is used to have a quick glance at the
generated binary code for \texttt{vulnerable\_function} and note down the
distance between the saved base pointer the start of the buffer too (line 17).

Putting this together yields following payload: Starting with some
\texttt{'A'}s to fill the buffer followed by 4 \texttt{'B'}s to overwrite the
saved base pointer. The next part is new, we attach the address of
\texttt{system} followed by some padding and a pointer to \texttt{not\_used}.

We happily receive a shell upon running the exploit. Execution will be handed
back to the original binary after we close the shell. Since we messed up the
control flow with our exploit the program segfaults shortly after.

This is also described as \textit{ret2libc} since we used ROP to jump to a
function (\texttt{system}) provided by libc.

@@

ASLR is enabled by default and one can easily check the implications by running
\texttt{cat} on \texttt{/proc/self/maps} a few times as shown in
\cref{src:aslr_cat}. Line 10, 19 and 28 show, for example, that the stack
starts at different locations in memory each time cat is invoked.

We can directly see one flaw in this setup --- not all sections of the
\texttt{cat} binary start at random locations. Especially the \texttt{.text}
always starts at the same position. This happens because \texttt{cat} itself
was not compiled as a Position Independent Executable (PIE). Since this is
actually the default of \texttt{gcc}, most programs' \texttt{.text} segment
will always start at the same location. One could pass the corresponding flag
(\texttt{-pie}) to the compiler to prevent this, so ASLR would be able to
randomize these segments too.

Breaking ASLR, even when the code is compiled with \texttt{-pie}, is easier
than it seems at first. Relocation only happens to a section at whole,
functions inside a section still share the same relative distance as they would
without ASLR. But before exploit this fact have a look at the randomized
addresses again.

Only three nybble ($3 \times \SI{4}{\bit}$) differ between multiple runs giving
us $2^{12} = 4096$ possibilities. If the scenario allows it, brute forcing is a
viable option here. But note that this changes drastically for \SI{64}{\bit}.
But we won't hassle with brute force, a better option has already been teased.

@@

ASLR can be broken easily as soon as \emph{one} pointer to a section of
interest gets \emph{leaked}. Therefore the name information leak. We show the
implications of such a leak by an example taken from \cite{rpisec}.

@@

In case you may wonder how easy it is to leak a pointer, this already happened
to us as a side effect in the format string example
(\cref{sec:format_string_exploit}).

Our previous exploit can be adapted as follows. First, manage to leak a pointer
somehow, which enables you to calculate the address offset introduced by ASLR.
Augment your ROP chain to take the offset into account. Run the exploit. Since
this is rather simple and we already gave an example how to calculate the
offset, we would like to leave this as an exercise for the reader.

Manipulate the target file used in the ROP example to print the address of
\texttt{printf} first, \emph{then} read in the payload via \texttt{stdin}. This
way you can first simulate a leaked pointer, adapt the ROP chain and run it.
Double check the distances between functions in the libraries, they may differ
with the ones used in this writeup.

@@

DEP can be fooled by ROP and ASLR is rendered useless with a simple info leak.
Something else is required at this point. Thinking back, the original problem
emerged from manipulating the return address located on the stack. Two
(additional) counter mechanisms were introduced going by the names StackGuard
and StackShield. We will take only a look at StackGuard and one relatively
common scenario to break it, but there is a comprehensive article
\cite{phrack_stack_guard} on \texttt{phrack}\footnote{\url{http://phrack.org/}}
describing and breaking both mechanisms.

The general idea behind StackGuard is to place \emph{something} before the
return address which \emph{guards} against overwriting the return address via a
buffer overflow. This \emph{something} is known as a canary and comes in
different forms.

\begin{description}
    \item[Terminator] A terminator canary contains a sequence of commonly used
        terminator symbols (like null, EOF, linefeed, \dots) to
        \emph{terminate} \emph{most} string operations before they would change
        the return address.

    \item[Random] A random canary is chosen at program start, stored
        \textit{somewhere save} and pushed into the stack upon function calls.
        The canary on the stack is checked against the one stored before
        executing the return instruction. The program is terminated upon
        mismatch. With this setup an attacker has to know the canary in order
        to overwrite the return address. Since it is chosen at random during
        program start, an attacker cannot reliably reproduce the same canary in
        his cloned setup.

@@

There is also the random XOR canary which XORs the (stored) random canary with
the return address before placing it on the stack. ``This is effectively
encryption of the return address with the random canary of this
function.''~\cite{phrack_stack_guard}

@@

As can be seen in \cref{src:canary_gcc} the buffer was not filled beyond its
capacity to examine the canary located in the same stack frame. A script
created by Daniel Walter\footnote{\url{http://0x90.at/post/gdb-stack-script}}
has been adapted slightly to display the stack together with some annotations.
Using \texttt{"AAAAAAA\textbackslash{}n"} and
\texttt{"BBBBBBB\textbackslash{}n"} makes the buffer clearly visible in lines
16 and 17. The canary can be observed in line 18.

The canary itself is composed of a terminator (null) followed by a random
sequence of \SI{3}{\byte}. This sequence changes very time the program is run.
Feeding more data to the buffer and overflowing it this way yields termination
of the program. Note that \texttt{puts} is still executed, the termination
happens just before the return of \texttt{fun}.

\begin{code*}{linenos=false}
    > echo AAAAAAAADEADBEEF | ./vuln
    AAAAAAAADEADBEEF

@@

\subsection{Server Worker Paradigm}

Of course there are multiple paths available when trying to break the
StackGuard mechanism, as mentioned in \cite{phrack_stack_guard}. We will now
have a look at the common server worker paradigm. \Cref{src:server_worker}
shows how that paradigm looks like from a task monitors view. The server /
daemon (here \texttt{apache2}) is started with root privileges in order to
listen on a \textit{privileged} port. After the initialization has been
compiled the server forks itself multiple times to create a set of workers. In
this example the workers drop their root privileges right away by changing
their current user to \texttt{www-data}. But our focus is not on the privileges
but the problem introduced by fork with respect to the StackGuard.

\begin{listing}[h!]
    \codefile[linenos=false]{gfx/server_worker.txt}
    \caption{Server worker paradigm from the view of a task monitor}
    \label{src:server_worker}
\end{listing}

Many things are copied\footnote{Actually referenced utilizing a copy-on-write
method} over to the new process when using \texttt{fork}. The canary is copied
too (more details at \cite{phrack_stack_guard2}). Together with the
fact\footnote{We assume that the server wants to maintain a maximum of
availability} that the server will fork itself again if one of its workers dies
or crashes to keep the worker pool at its configured sized.

An attacker will be able to guess the \emph{same} canary multiple times since
the server will keep spawning workers if they crash --- even due to a stack
smash. The attacker receives information about whether his guess was correct or
not by whether his connection has been terminated. And now to the meat of this
method.

Have a look at \cref{src:canary_gcc} again and reexamine the canary. While
occupying \SI{4}{\byte} only 3 of them are random --- first byte acts as a
terminator. We have already seen via previous examples that a buffer overflow
often allows writing to consecutive memory \emph{byte by byte}. Putting this
information together yields following upper bound for brute forcing a canary in
the described scenario:

\begin{align*}
    \implies 2^8 \times 3 & = 768  & \text{ guesses at most on \SI{32}{\bit}}\\
    \implies 2^8 \times 7 & = 1792 & \text{ guesses at most on \SI{64}{\bit}}
\end{align*}

Again, this is just \emph{one} of many different ways to work around the
StackGuard mechanism. Depending on your operating system's and compiler's
implementation this may or may not work. We encourage the reader to try this
technique locally with a minimal example. Running the exploit multiple times
and recording the runtime (number of guesses) may be of interest.
